Using Distributed Information Stores for Patient-Clinician Health Data Sharing

### 2013-01-16 patient-clinician-workshop/pccworkshop.tex
Spelling

product & measurement indices & price range (approx. USD) \\
\hline
FitBit & general activity (walking/jogging) & \$50 \\
Nike FuelBand & general activity (walking/jogging) & \$199 \\ 
WiThings WiFi Scale & weight & \$200 \\
WiThings BPM & blood pressure & \$200 \\ 
Zeo Pro & sleep duration \& quality & \$200 \\
WakeMate & sleep duration & \$50  \\
Zensorium Tinnke & heart rate variability, blood oxygenation, respiratory rate & \$200 \\
BodyMedia CORE 2 & galvanic skin response, body temperature, sleep, respiratory rate, heart rate & (TBD) \\
Hapifork & eating time, rate, frequency & (TBD) \\ 
\end{tabular}
\caption{\emph{Consumer-grade activity sensors}: Activity and physiological measurement sensors with automatic data uplink to cloud services }
\label{sensors}
\end{table}
}
\vspace{6cm}

@@

While few of these devices had initially been evaluated for accuracy
or certified for clinical use, increasingly device manufacturers have
begun to get independent certification. For example, BodyMedia
advertises its device as ``FDA-approved, clincally validated, most
accurate wireless activity measurement''.  A separate study of
activity-based sleep monitoring tools revealed that ``low cost
actigraphy-based approaches'' correlated well with baseline (in-lab
EEG-based measurements) for sleep duration, although less with sleep
quality metrics \cite{reid1999correlation}.

@@

\section{Storing and Representing Activity Diaries}
Currently, these devices are designed to upload measurements only to separate, dedicated web sites run by their manufacturers.  Such sites are poorly suited for the long-term storage of people's activity data for several reasons.  First, none of the sites providelong-term retention or access guarantees of user data (nor HIPAA security guarantees!).  Second, it is difficult for a user to gain a unified understanding of their activity through the disparate interfaces provided, in part due of the widely different visual and interface representations. Finally, many of interfaces represent activity in arbitrary, ``user-friendly'' units, such as ``Sleep score'', ``Fuel'', ``Zen level'' etc., which are difficult or impossible for users to compare.

However, most of these sites have APIs providing programmatic access to raw sensor data.  Our approach has been to leverage such APIs to retrieve a patient's information from all their devices, consolidating it into a uniform, common taxonomy and description logic, like those provided by SNOMED-CT \cite{stearns2001snomed} and ICD-10 \cite{world1993icd}.  The goal of such a formalisation is to provide a standard descriptive mechanism by which this data could be interpreted, such as by future clinical medical systems.

Since SNOWMED-CT nor ICD cover activity-related concepts, we devised a supplemental ontology called \emph{Activity-SENSE}.  This ontology encompasses patient activities, and standard units of measure with which activities measurements were taken, and concepts for describing sensors used.  The intention is for records in Activity-SENSE, like those of SNOMED-CT, will be eventually translatable to dozens of different languages and mapped to similar activity ontologies, allowing international use and greater operability.

>>>>

product & measurement indices & price range (approx. USD) \\
\hline
FitBit & general activity (walking/jogging) & \$50 \\
Nike FuelBand & general activity (walking/jogging) & \$199 \\ 
WiThings WiFi Scale & weight & \$200 \\
WiThings BPM & blood pressure & \$200 \\ 
Zeo Pro & sleep duration \& quality & \$200 \\
WakeMate & sleep duration & \$50  \\
Zensorium Tink\'{e} & heart rate variability, blood oxygenation, respiratory rate & \$200 \\
BodyMedia CORE 2 & galvanic skin response, body temperature, sleep, respiratory rate, heart rate & (TBD) \\
Hapifork & eating time, rate, frequency & (TBD) \\ 
\end{tabular}
\caption{\emph{Consumer-grade activity sensors}: Activity and physiological measurement sensors with automatic data uplink to cloud services }
\label{sensors}
\end{table}
}
\vspace{6cm}

@@

While few of these devices had initially been evaluated for accuracy
or certified for clinical use, increasingly device manufacturers have
begun to get independent certification. For example, BodyMedia
advertises its device as ``FDA-approved, clinically validated, most
accurate wireless activity measurement''.  A separate study of
activity-based sleep monitoring tools revealed that ``low cost
actigraphy-based approaches'' correlated well with baseline (in-lab
EEG-based measurements) for sleep duration, although less with sleep
quality metrics \cite{reid1999correlation}.

@@

\section{Storing and Representing Activity Diaries}
Currently, these devices are designed to upload measurements only to separate, dedicated web sites set up by their manufacturer.  Such sites are poorly suited for the long-term storage of people's activity data for several reasons.  First, none of the sites provide long-term retention or access guarantees of user data (nor HIPAA security guarantees!).  Second, it is difficult for a user to gain a unified understanding of their activity through the disparate interfaces provided, in part due of the widely different visual and interface representations. Finally, many of interfaces represent activity in arbitrary, ``user-friendly'' units, such as ``Sleep score'', ``Fuel'', ``Zen level'' etc., which are difficult or impossible for users to compare.

However, most of these sites have APIs providing programmatic access to raw sensor data.  Our approach has been to leverage such APIs to retrieve a patient's information from all their devices, consolidating it into a uniform, common taxonomy and description logic, like those provided by SNOMED-CT~\cite{stearns2001snomed} and ICD-10~\cite{world1993icd}.  A uniform representation enables such data to be interpretable and used directly by future clinical medical systems, such as primary care information management systems.

Since SNOMED-CT nor ICD cover activity-related concepts, we devised a supplemental ontology called \emph{Activity-SENSE}.  This ontology encompasses patient activities, and standard units of measure with which activities measurements were taken, and concepts for describing sensors used.  The intention is for records in Activity-SENSE, like those of SNOMED-CT, will be eventually translatable to dozens of different languages and mapped to similar activity ontologies, allowing international use and greater operability.
